Elaboration with First-Class Implicit Function Types

### 2020-02-29 paper.tex
stuff

\newcommand{\echeckblank}{\llbracket\blank\rrbracket\!\Downarrow}
\newcommand{\einferblank}{\llbracket\blank\rrbracket\!\Uparrow}
\newcommand{\echeck}[4]{\llbracket#1\rrbracket\!\Downarrow\,_{#2|#3}\,#4}
\newcommand{\einfer}[4]{\llbracket#1\rrbracket\!\Uparrow\,_{#3|#4}\,#2}
\newcommand{\echeckt}[2]{\llbracket#1\rrbracket\!\Downarrow\,#2}
\newcommand{\einfert}[2]{\llbracket#1\rrbracket\!\Uparrow\,#2}
\newcommand{\edo}{\boldsymbol{\mathsf{do}}}
\newcommand{\ereturn}{\boldsymbol{\mathsf{return}}}
\newcommand{\eif}{\boldsymbol{\mathsf{if}}}
\newcommand{\ethen}{\boldsymbol{\mathsf{then}}}
\newcommand{\eelse}{\boldsymbol{\mathsf{else}}}
\newcommand{\ecase}{\boldsymbol{\mathsf{case}}}
\newcommand{\eof}{\boldsymbol{\mathsf{of}}}
\newcommand{\true}{\mathsf{true}}
\newcommand{\false}{\mathsf{false}}
\newcommand{\einsert}[3]{\mathsf{insert}\,#1\,#2\,#3}
\newcommand{\einsertpar}[5]{\mathsf{insert}_{#1|#2}\,#3\,#4\,#5}
\newcommand{\Set}{\mathsf{Set}}
\renewcommand{\tt}{\mathsf{tt}}

@@

\begin{definition}[Contextualization]\label{def:contextualization} For each $\Theta|\Gamma \vdash t : \Gamma\To A$,
we define the spine $\overline{\mathsf{vars}_{\Gamma}}$ such that that
$\Theta|\Gamma\vdash t\,\overline{\mathsf{vars}_{\Gamma}} : A$. Informally, this
is $t$ applied to all bound variables in $\Gamma$.
  \begin{alignat*}{3}
    &(t\,\overline{\mathsf{vars}_{\Gamma,\,x:A}})   && :\equiv (t\,\overline{\mathsf{vars}_{\Gamma}})\,x \\
    &(t\,\overline{\mathsf{vars}_{\Gamma,\,x:A=t}}) && :\equiv (t\,\overline{\mathsf{vars}_{\Gamma}}) \\
    &(t\,\overline{\mathsf{vars}_{\emptycon}})     && :\equiv t
  \end{alignat*}
\end{definition}

@@

\subsection{Implicit Argument Insertion}

We define a helper function which inserts implicit applications around a core
term. For example, if we have a defined name $id$ with type $\{A : \U\}\to A \to
A$ in a surface program, we usually want to expand $id$ to $id\,\{\alpha\}$,
where $\alpha$ is a fresh metavariable.
\begin{mathpar}
\inferrule*{ins \in \{\true,\,\false\} \\ \Theta|\Gamma\vdash t : A}
           {\einsert{ins}{t}{A} \in \{(\Theta',\,\theta\,t',\,A')\,|\,
             (\theta : \Theta' \To \Theta)\,\land\,(\Theta'|\Gamma[\theta]\vdash t' : A')\}}
\end{mathpar}
We have an additional $ins \in \{\true,\,\false\}$ parameter, which simply
toggles whether any insertion is to be performed. We will use this in the
definition of elaboration. Insertion is defined by recursion on $ins$ and the
$A$ type, as follows. Here, we notate $\Theta$ and $\Gamma$ as subscripts but
later we will leave them implicit. The clauses of the following definition are
matched top-down.
\begin{alignat*}{3}
  &\einsertpar{\Theta}{\Gamma}{\false}{t}{A} && :\equiv\,\,&& (\Theta,\,\id,\,t,\,A)\\
  &\einsertpar{\Theta}{\Gamma}{\true\,}{t}{(\{x : A\}\to B)} && :\equiv&&
       \slet\,(\Theta',\,\theta,\,u) = \freshMeta{\Theta}{\Gamma}{A}\\
  & && &&\sin\,\einsertpar{\Theta'}{\Gamma[\theta']}{\true}{((t[\theta])\,\{u\})}{(B[\theta][x\mapsto u])}\\
  &\einsertpar{\Theta}{\Gamma}{\true\,}{t}{A} && :\equiv&& (\Theta,\,\id,\,t,\,A)
\end{alignat*}

@@

In this section we define the elaboration algorithm. First, we explain the used notations.

\begin{itemize}
  \item We use a Haskell-like monadic pseudocode notation, where the side effect is
    failure via $\fail$.
  \item We use pattern matching notation on core terms; e.g.\ we may match on
    whether a type is a function type. This assumes an evaluation/normalization
    procedure on core terms; but note that we already assume this in implicit
    argument insertion and unification.
  \item We abbreviate $\theta_1 \circ \theta_2$ as $\theta_{12}$, $\theta_1
    \circ \theta_2 \circ \theta_3$ as $\theta_{123}$ and analogously in other
    cases. We do this to reduce the visual noise caused by threading composed
    metasubstitutions everywhere in the elaboration algorithm.
\end{itemize}

Elaboration consists of two (partial) functions, checking and inferring, which
are defined by mutual induction on surface syntax. They are specified as
follows. We also have the additional $ins \in \{\true,\,\false\}$ argument for
inference, which toggles whether we perform implicit argument insertion on the
output. We explain the cases of the definition in order.

\begin{mathpar}
\inferrule*[lab=check]
           {\text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash A : \U}
           {\echeck{t}{\Theta}{\Gamma}{A} \in
             \{(\Theta',\,\theta,\,t')\,|\,(\theta : \Theta' \To \Theta)\,\land\,
               (\Theta'|\Gamma[\theta]\vdash t' : A[\theta])\}\cup\{\fail\}
           }

\inferrule*[lab=infer]
           {ins \in \{\true,\,\false\} \\
                \text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash}
           {\einfer{t}{ins}{\Theta}{\Gamma} \in
             \{(\Theta',\,\theta,\,t',\,A)\,|\,(\theta : \Theta' \To \Theta)\,\land\,
               (\Theta'|\Gamma[\theta]\vdash t' : A)\}\cup\{\fail\}
           }
\end{mathpar}

\begingroup
\allowdisplaybreaks

@@

  &\quad(\Theta_1,\,\theta_1,\,t',\,B) \leftarrow \einfer{t}{\true}{\Theta_0}{\Gamma}\\

@@

  &\einfer{x}{ins}{\Theta}{\Gamma} :\equiv \edo\\

@@

  &\qquad   \ethen\, \ereturn\,(\einsert{ins}{x}{A})\\

@@

  &\einfer{\U}{ins}{\Theta}{\Gamma} :\equiv \edo\\

@@

  &\einfer{(x : A)\to B}{ins}{\Theta_0}{\Gamma} :\equiv \edo\\

@@

  &\einfer{\{x : A\}\to B}{ins}{\Theta_0}{\Gamma} :\equiv \edo\\

@@

  &\einfer{\lambda\,x.\,t}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
  &\quad(\Theta_1,\,\theta_1,\,A)\leftarrow \freshMeta{\Theta_0}{\Gamma}{\U}\\
  &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \einfer{t}{\true}{\Theta_1}{\Gamma[\theta_1],\,x:A}\\

@@

  &\einfer{\lambda\,\{x\}.\,t}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
  &\quad(\Theta_1,\,\theta_1,\,A)\leftarrow \freshMeta{\Theta_0}{\Gamma}{\U}\\
  &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \einfer{t}{\true}{\Theta_1}{\Gamma[\theta_1],\,x:A}\\
  &\quad\ereturn\,(\einsert{ins}{(\lambda\,\{x\}.\,t')}{(\{x : A[\theta_2]\}\to B)})\\
  &\einfer{t\,u}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
  &\quad (\Theta_1,\,\theta_1,\,t',\,A) \leftarrow \einfer{t}{\true}{\Theta_0}{\Gamma}\\
  &\quad (\Theta_2,\,\theta_2,A_0) \leftarrow \freshMeta{\Theta_1}{\Gamma[\theta_1]}{\U}\\
  &\quad (\Theta_3,\,\theta_3,A_1) \leftarrow \freshMeta{\Theta_2}{\Gamma[\theta_{12}],\,x:A_0}{\U}\\

@@

  &\quad \ereturn\,(\mathsf{insert}\,{ins}\,{((t'[\theta_{2345}])\,u')}\,{(A_1[\theta_{45}][x\mapsto u'])})\\
  &\einfer{t\,\{u\}}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
  &\quad (\Theta_1,\,\theta_1,\,t',\,A) \leftarrow \einfer{t}{\false}{\Theta_0}{\Gamma}\\
  &\quad (\Theta_2,\,\theta_2,A_0) \leftarrow \freshMeta{\Theta_1}{\Gamma[\theta_1]}{\U}\\
  &\quad (\Theta_3,\,\theta_3,A_1) \leftarrow \freshMeta{\Theta_2}{\Gamma[\theta_{12}],\,x:A_0}{\U}\\

@@

  &\quad \ereturn\,(\mathsf{insert}\,{ins}\,{((t'[\theta_{2345}])\,\{u'\})}\,{(A_1[\theta_{45}][x\mapsto u'])})\\
  &\einfer{\slet\,x:A=t\,\sin\,u}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\

@@

                \einfer{u}{ins}{\Theta_2}{\Gamma[\theta_{12}]}\\

@@

  &\einfer{\_}{ins}{\Theta}{\Gamma} :\equiv \edo \\

@@

@@

We provide an implementation of the elaborator from Section
\ref{sec:extending_elaboration}. It is a standalone Haskell program which reads
a surface expression from standard input, and outputs the result of elaboration,
or optionally the type or the normal form of the result. It is implemented in
about 1200 lines of Haskell. Of this, 673 lines constitute the core syntax,
evaluation, unification and elaboration. Of this 673 lines, 458 lines implement
the basic elaborator of Section \ref{sec:bidirectional_elaboration} and 215
lines implement the extended elaborator.

>>>>

\newcommand{\echeckblank}{\llbracket\blank\rrbracket\!\Downarrow}
\newcommand{\einferblank}{\llbracket\blank\rrbracket\!\Uparrow}
\newcommand{\echeck}[4]{\llbracket#1\rrbracket\!\Downarrow\,_{#2|#3}\,#4}
\newcommand{\einfer}[3]{\llbracket#1\rrbracket\!\Uparrow\,_{#2|#3}}
\newcommand{\echeckt}[2]{\llbracket#1\rrbracket\!\Downarrow\,#2}
\newcommand{\einfert}[2]{\llbracket#1\rrbracket\!\Uparrow\,#2}
\newcommand{\edo}{\boldsymbol{\mathsf{do}}}
\newcommand{\ereturn}{\boldsymbol{\mathsf{return}}}
\newcommand{\eif}{\boldsymbol{\mathsf{if}}}
\newcommand{\ethen}{\boldsymbol{\mathsf{then}}}
\newcommand{\eelse}{\boldsymbol{\mathsf{else}}}
\newcommand{\ecase}{\boldsymbol{\mathsf{case}}}
\newcommand{\eof}{\boldsymbol{\mathsf{of}}}
\newcommand{\true}{\mathsf{true}}
\newcommand{\false}{\mathsf{false}}
\newcommand{\einsert}{\mathsf{insert}}
\newcommand{\Set}{\mathsf{Set}}
\renewcommand{\tt}{\mathsf{tt}}

@@

\begin{definition}[Contextualization]\label{def:contextualization}
For each $\Theta|\Gamma \vdash t : \Gamma\To A$, we define the spine
$\overline{\mathsf{vars}_{\Gamma}}$ such that that $\Theta|\Gamma\vdash
t\,\overline{\mathsf{vars}_{\Gamma}} : A$, which is $t$ applied to all bound
variables in $\Gamma$.
  \begin{alignat*}{3}
    &(t\,\overline{\mathsf{vars}_{\Gamma,\,x:A}})   && :\equiv (t\,\overline{\mathsf{vars}_{\Gamma}})\,x \\
    &(t\,\overline{\mathsf{vars}_{\Gamma,\,x:A=t}}) && :\equiv (t\,\overline{\mathsf{vars}_{\Gamma}}) \\
    &(t\,\overline{\mathsf{vars}_{\emptycon}})     && :\equiv t
  \end{alignat*}
\end{definition}

@@

@@

Elaboration consists of two (partial) functions, checking and inferring, which
are defined by mutual induction on surface syntax. We also have implicit
argument insertion as a helper function, which is defined by recursion on core
types, and which is used as a post-processing step after inference. First, about
the used notations:
\begin{itemize}
  \item We use a Haskell-like monadic pseudocode notation, where the side effect is
    failure via $\fail$.
  \item We use pattern matching notation on core terms; e.g.\ we may match on
    whether a type is a function type. This assumes an evaluation/normalization
    procedure on core terms; but note that we already assume this in
    unification.
  \item We abbreviate $\theta_1 \circ \theta_2$ as $\theta_{12}$, $\theta_1
    \circ \theta_2 \circ \theta_3$ as $\theta_{123}$ and analogously in other
    cases. We do this to reduce the visual noise caused by threading composed
    metasubstitutions everywhere in the elaboration algorithm.
\end{itemize}
We present the specifications and definitions below, then we describe them in order.
\begin{mathpar}
\inferrule*[lab=insert]{\Theta_0|\Gamma\vdash \\ (\Theta,\,\theta,\,t,\,A) \in
             \{(\Theta,\,\theta,\,t,\,A)\,|\,(\theta : \Theta_0 \To \Theta)\,\land\,
               (\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}}
           {\einsert(\Theta,\,\theta,\,t,\,A) \in \{(\Theta,\,\theta,\,t,\,A)\,|\,
             (\theta : \Theta_0 \To \Theta)\,\land\,(\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}}
\end{mathpar}
\begin{alignat*}{3}
  &\einsert\,(\Theta_1,\,\theta_1,\,t,\,\{x : A\}\to B) && :\equiv\hspace{0.5em}&&
       \slet\,(\Theta_2,\,\theta_2,\,u) = \freshMeta{\Theta}{\Gamma}{A}\\
  & && &&\sin\,\einsert\,(\Theta_2,\,\theta_{12},\,(t[\theta_2])\,\{u\},\,B[\theta_2][x\mapsto u])\\
  &\einsert\,(\Theta_1,\,\theta_1,\,t,\,A) && :\equiv&& \ereturn\,(\Theta_1,\,\theta_1,\,t,\,A)\\
  &\einsert\,\fail && :\equiv && \fail
\end{alignat*}
%% \pagebreak
\begingroup
\allowdisplaybreaks
\begin{mathpar}
\inferrule*[lab=check]
           {\text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash A : \U}
           {\echeck{t}{\Theta}{\Gamma}{A} \in
             \{(\Theta',\,\theta,\,t')\,|\,(\theta : \Theta' \To \Theta)\,\land\,
               (\Theta'|\Gamma[\theta]\vdash t' : A[\theta])\}\cup\{\fail\}
           }

\inferrule*[lab=infer]
           {\text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash}
           {\einfer{t}{\Theta}{\Gamma}\,\in
             \{(\Theta',\,\theta,\,t',\,A)\,|\,(\theta : \Theta' \To \Theta)\,\land\,
               (\Theta'|\Gamma[\theta]\vdash t' : A)\}\cup\{\fail\}
           }
\end{mathpar}

@@

  &\quad(\Theta_1,\,\theta_1,\,t',\,B) \leftarrow \einsert\,(\einfer{t}{\Theta_0}{\Gamma})\\

@@

  &\einfer{x}{\Theta}{\Gamma} :\equiv \edo\\

@@

  &\qquad   \ethen\, \ereturn\,(\Theta,\,\id,\,x,\,A)\\

@@

  &\einfer{\U}{\Theta}{\Gamma} :\equiv \edo\\

@@

  &\einfer{(x : A)\to B}{\Theta_0}{\Gamma} :\equiv \edo\\

@@

  &\einfer{\{x : A\}\to B}{\Theta_0}{\Gamma} :\equiv \edo\\

@@

  &\einfer{\lambda\,x.\,t}{\Theta_0}{\Gamma} :\equiv \edo \\
  &\quad \slet\, (\Theta_1,\,\theta_1,\,A) = \freshMeta{\Theta_0}{\Gamma}{\U}\\
  &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \einsert\,(\einfer{t}{\Theta_1}{\Gamma[\theta_1],\,x:A})\\

@@

  &\einfer{\lambda\,\{x\}.\,t}{\Theta_0}{\Gamma} :\equiv \edo \\
  &\quad \slet\,(\Theta_1,\,\theta_1,\,A) = \freshMeta{\Theta_0}{\Gamma}{\U}\\
  &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \einsert\,(\einfer{t}{\Theta_1}{\Gamma[\theta_1],\,x:A})\\
  &\quad\ereturn\,(\Theta_2,\,\theta_{12},\,\lambda\,\{x\}.\,t',\,\{x : A[\theta_2]\}\to B)\\
  &\einfer{t\,u}{\Theta_0}{\Gamma} :\equiv \edo \\
  &\quad (\Theta_1,\,\theta_1,\,t',\,A) \leftarrow \einsert\,(\einfer{t}{\Theta_0}{\Gamma})\\
  &\quad \slet\,(\Theta_2,\,\theta_2,A_0) = \freshMeta{\Theta_1}{\Gamma[\theta_1]}{\U}\\
  &\quad \slet\,(\Theta_3,\,\theta_3,A_1) = \freshMeta{\Theta_2}{\Gamma[\theta_{12}],\,x:A_0}{\U}\\

@@

  &\quad \ereturn\,(\Theta_5,\,\theta_{12345},\,(t'[\theta_{2345}])\,u',\,A_1[\theta_{45}][x\mapsto u'])\\
  &\einfer{t\,\{u\}}{\Theta_0}{\Gamma} :\equiv \edo \\
  &\quad (\Theta_1,\,\theta_1,\,t',\,A) \leftarrow \einfer{t}{\Theta_0}{\Gamma}\\
  &\quad \slet\,(\Theta_2,\,\theta_2,A_0) = \freshMeta{\Theta_1}{\Gamma[\theta_1]}{\U}\\
  &\quad \slet\,(\Theta_3,\,\theta_3,A_1) = \freshMeta{\Theta_2}{\Gamma[\theta_{12}],\,x:A_0}{\U}\\

@@

  &\quad \ereturn\,(\Theta_5,\,\theta_{12345},\,(t'[\theta_{2345}])\,\{u'\},\,A_1[\theta_{45}][x\mapsto u'])\\
  &\einfer{\slet\,x:A=t\,\sin\,u}{\Theta_0}{\Gamma} :\equiv \edo \\

@@

                \einfer{u}{\Theta_2}{\Gamma[\theta_{12}]}\\

@@

  &\einfer{\_}{\Theta}{\Gamma} :\equiv \edo \\

@@

\subsubsection{Implicit argument insertion}
This inserts implicit applications around a core term. For example, if we have a
defined name $id$ with type $\{A : \U\}\to A \to A$ in a surface program, we
usually want to expand $id$ to $id\,\{\alpha\}$, where $\alpha$ is a fresh
metavariable. We define $\einsert$ to take as input the output of
$\einferblank$, so that it is more convenient to use as an optional
post-processing step after inference.

@@

We provide an implementation of the elaborator from Section
\ref{sec:extending_elaboration}. It is a standalone Haskell program which reads
a surface expression from standard input, and outputs the result of elaboration,
or optionally the type or the normal form of the result. It is implemented in
1061 lines of Haskell. Of this, 673 lines constitute the core syntax,
evaluation, unification and elaboration. Of these 673 lines, 458 lines implement
the basic elaborator of Section \ref{sec:bidirectional_elaboration} and 215
lines implement the extended elaborator.
